---
title: "StormData"
author: "Andriy Levytskyy"
date: "18/09/2020"
output: html_document
---
# Relation of Severe Weather Events on Public Health and Economy

## Synopsis
It was found that severe weather events indeed had a huge impact on society in recent years. Floods were found to have cost most significant economy damage, which attributed to more than 150 billions US dollars of property damage. Tornados were found to have made most number of death and injuries, with almost 97,000 injuries or fatalities in recent years.

## Data Processing

**Assumptions**:
* The file has been download to the current working directory
* The file was named StormData.csv.bz2

**Performed steps**:
* Load the data from bzipped file
* Calculate the damage
 * Using Logic, it was found that the exponential has lots of values that were not explained in the documentation. These values were ignored. Only K, M, B were understood as 1000, 10^6 and 10^9, respectively.
 * Property damage & crop damage were summed up to get the total damage
* Calculate total fatalities and injuries
* Plot top 10 event

```{r}
# Load Libraries
library(dplyr)
library(ggplot2)

filename<-"StormData.csv.bz2"

# Download file

if(!file.exists(filename)){
  fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(fileURL, filename, method = "curl")
}

# Read data 

StormData<-read.csv("StormData.csv.bz2", header = TRUE)

StormData_by_evtype<-group_by(StormData, EVTYPE)

StormData_sum_by_evtype<-summarise(StormData_by_evtype,  
                                  SumFATALITIES=sum(FATALITIES), SumINJURIES=sum(INJURIES), 
                                  TotalHealth=SumFATALITIES+SumINJURIES)

StormData_sum_by_evtype<-StormData_sum_by_evtype[order(StormData_sum_by_evtype$TotalHealth, decreasing = TRUE), ]

StormData_sum_by_evtype_top10<-head(StormData_sum_by_evtype, 10)

```

## Results
### Impact on Population health
Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r}
# top 10 by Total health damage
StormData_sum_by_evtype_top10

# Plot top 10
ggplot(StormData_sum_by_evtype_top10, aes(x=EVTYPE, y=TotalHealth, fill=EVTYPE)) +
  geom_bar(stat="identity") +
  ggtitle("Top 10 casualties due to severe weather events in US") +
  xlab("Events") +
  ylab("Total casualties") +
  theme(axis.text.x = element_text(angle = 90, vjust=0.5))

```


From the figure, it was found that Tornado has caused the most number of injuries and fatalities (96,979 fatalities and injuries), significantly more than any other type of events.

### Impact on Economy

```{r}
StormData_by_evtype<-mutate(StormData_by_evtype, PropDamage = ifelse(PROPDMGEXP == "", PROPDMG, 0 ))

StormData_by_evtype$PropDamage<-with(StormData_by_evtype, 
                                     ifelse(PROPDMGEXP == "K", PROPDMG*1000+PropDamage, PropDamage))

StormData_by_evtype$PropDamage<-with(StormData_by_evtype, 
                                     ifelse(PROPDMGEXP == "M", PROPDMG*1000000+PropDamage, PropDamage ))

StormData_by_evtype$PropDamage<-with(StormData_by_evtype, 
                                     ifelse(PROPDMGEXP == "B", PROPDMG*1000000000+PropDamage, PropDamage ))


StormData_by_evtype<-mutate(StormData_by_evtype, CropDamage = ifelse(CROPDMGEXP == "", CROPDMG, 0 ))

StormData_by_evtype$CropDamage<-with(StormData_by_evtype, 
                                     ifelse(CROPDMGEXP == "K", CROPDMG*1000+CropDamage, CropDamage))

StormData_by_evtype$CropDamage<-with(StormData_by_evtype, 
                                     ifelse(CROPDMGEXP == "M", CROPDMG*1000000+CropDamage, CropDamage ))

StormData_by_evtype$CropDamage<-with(StormData_by_evtype, 
                                     ifelse(CROPDMGEXP == "B", CROPDMG*1000000000+CropDamage, CropDamage ))


StormData_sum_damage_by_evtype<-summarise(StormData_by_evtype,  
                                   SumPropDamage=sum(PropDamage), SumCropDamage=sum(CropDamage),
                                   TotalDamage=SumPropDamage+SumCropDamage)

StormData_sum_damage_by_evtype<-StormData_sum_damage_by_evtype[order(StormData_sum_damage_by_evtype$TotalDamage, decreasing = TRUE), ]


StormData_sum_damage_by_evtype_top10<-head(StormData_sum_damage_by_evtype, 10)

StormData_sum_damage_by_evtype_top10

ggplot(StormData_sum_damage_by_evtype_top10, aes(x=EVTYPE, y=TotalDamage, fill=EVTYPE)) +
  geom_bar(stat="identity") +
  ggtitle("Top 10 weather economic consequences events in US") +
  xlab("Events") +
  ylab("Total damage in millions USD") +
  theme(axis.text.x = element_text(angle = 90, vjust=0.5))

```

From the figure, it was found that Flood has caused biggest damage (around 150 billions US dollars), much higher than any other events.